---
title: "Exercise 2"
author: "Hubert Rehrauer, modified by Cyril Statzer"
date: "25 9 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Exploratory Data Analysis

Do an exploratory data analysis of a matrix of expression values. Load the data and display:

```{r}
#install.packages("limma")
library(limma)
#install.packages("pheatmap")
library(pheatmap)
```

## Data Import
```{r }
anno = read.table("SampleAnnotation.txt", as.is=TRUE, sep="\t", quote="",
                  row.names=1, header=TRUE)
x = read.table("expressiondata.txt", as.is=TRUE, sep="\t", quote="", row.names=1, header=TRUE, check.names = FALSE)
x = log2(as.matrix(x))
```

## Define samples and colors and phenotype
```{r}
samples = rownames(anno)
colors = rainbow(nrow(anno))
isNorm = anno$TissueType == "norm"
isSick = anno$TissueType == "sick"
isAcute = anno$TissueType == "acute"
```

Distribution analysis
*boxplot*
```{r boxplot}
boxplot(x,use.cols=1)
```

*density*
```{r densities}
limma::plotDensities(x,main = "Densities",legend=F,col = terrain.colors(nrow(anno)))
legend("topright",legend = colnames(x),cex = 0.7,col=terrain.colors(nrow(anno)),lty = 1, lwd = 4, y.intersp = 0.8)
```

Principle component analysis
```{r}
pca <- prcomp(x, center = T, scale. = T)
plot(pca, main = "PCA")
```

```{r}
#install.packages("ggbiplot")
#library(ggbiplot)
#g <- ggbiplot(pca)
```

Build correlation matrix from expression matrix (corr(x))
Normalization is performed using the min-max methond
```{r}
corr <- cor(x)
corr <- (corr - min(corr))/(max(corr) - min(corr))
#corr <- normalizeQuantiles(corr)
```


Generate a heatmap to analyze clustering of samples
Both the heatmap and pheatmap functions were used.
The general clustering is very good, indicating a strong difference between normal and other samples (acute and disease). Within the acute and disease groups the clustering is very weak and they cannot be separated.
```{r }
heatmap(corr,main = "Correlation of disease states")
pheatmap(corr,main = "Correlation of disease states")
```

Similarly to the heatmap here the clustering between patients is visualized in a separate dendrogram. The sick-14 sample is in the other cluster. All other acute and sick samples cluster within the same cluster.

* clustering: *hclust*
```{r}
hc <- hclust(dist(corr))
plot(hc)
```




